Using Optimal Time Step Selection to Boost the Accuracy of Finite Difference Methods for a Class of Time Dependent Partial Differential Equations

### 2008-05-14 papers/OTSPDE/OTSPDE.tex
OTSPDE: renamed LaTeX files from optimal_time_steps_for_pdes to OTSPDE.
OTSPDE:  added references to work on the use of spectral methods for solving problems on irregular domains.

\subsection{Boundary Conditions}
In our analysis and discussion thus far, we have completely ignored the issue 
of boundary conditions.  The choice of discretization for boundary conditions
can have a significant impact on the accuracy of the numerical solution.  In 
general, high-order discretizations of boundary conditions are required to 
ensure that errors introduced at the domain boundary do not pollute the 
numerical solution in the domain interior.  For Dirichlet boundaries, it
is sufficient to impose boundary conditions by using boundary (or ghost cell) 
values that have the same order of accuracy as the solution.  
More effort is required to ensure that Neumann boundary conditions do not
reduce the order of accuracy of the solution.  In general, Neumann boundary
conditions require that the accuracy of ghost cells slightly exceed the 
desired order of accuracy for the solution.  A detailed discussion of 
appropriate methods for imposing boundary conditions is beyond the scope of 
this paper.  We refer interested readers to the thorough analysis presented
in~\cite{gko_book}.  To avoid complicating the discussion in 
Sections~\ref{sec:applications_1d} and~\ref{sec:applications_multidim}, 
Neumann boundary conditions are imposed by taking advantage of knowledge of 
the analytical solution.

@@

\subsubsection{Irregular Domains}
All of the previous examples share an important feature: they are all
defined on simple, rectangular domains.  Thus, any of several methods 
could have been used to obtain solutions with similar or even higher accuracy.
For example, a fourth-order accurate solution to the diffusion equation
(in any number of space dimensions) could be obtained by using a 
pseudospectral spatial discretization~\cite{trefethen_spectral_book, 
boyd_spectral_book} of the Laplacian and a fourth-order time integration scheme 
(\eg Runge-Kutta).  When the domain is irregularly shaped, high-order methods 
are harder to design and implement.  For instance, spectral and pseudospectral 
methods are not usually appropriate because the domain is not a simple 
cartesian product of intervals.

@@

@@

\begin{figure}[tb]
\begin{center}
\scalebox{0.25}{\includegraphics{figures/ghostcell_edge}} 
\ \ \ \ \ \ \ \ \ \ \ \ \ 
\scalebox{0.25}{\includegraphics{figures/ghostcell_corner}}
\caption{Illustrations of edge (left) and corner (right) ghostcells.
The edge ghostcell, $u_E$, is filled by using cubic extrapolation of 
the value on the boundary, $u_B$, and values at three interior grid points.  
The corner ghostcell, $u_C$, is filled by using a fourth-order accurate
Taylor series expansion centered at $u_{0,0}$, which uses the values of $u$ 
at the specified interior grid points and neighboring edge ghost cells to 
approximate the partial derivatives in the expansion. 
}
\label{fig:ghost_cells}
\end{center}
\end{figure}

@@

\begin{figure}[tb]
\begin{center}
\scalebox{0.35}{
  \includegraphics{figures/diffusion_eqn_2d_starfish_domain_error_vs_N}} 
\caption{$L^\infty$ error as a function of number of grid points for two
finite difference schemes that solve the 2D diffusion equation on an 
irregular domain: forward Euler with optimal time step and correction 
terms (circles) and forward Euler with suboptimal time step $\dt = \dx^2/8 D$
(squares).  Both methods solve the 2D diffusion equation described in the 
caption for Figure~\ref{fig:diffusion_eqn_2d_starfish_domain}.
}
\label{fig:diffusion_eqn_2d_starfish_error}
\end{center}
\end{figure}

@@

\bibliography{optimal_time_steps_for_pdes}

>>>>

\subsection{Boundary Conditions}
In our analysis and discussion thus far, we have completely ignored the issue 
of boundary conditions.  The choice of discretization for boundary conditions
can have a significant impact on the accuracy of the numerical solution.  In 
general, high-order discretizations of boundary conditions are required to 
ensure that errors introduced at the domain boundary do not pollute the 
numerical solution in the domain interior.  For Dirichlet boundaries, it
is sufficient to impose boundary conditions by using boundary (or ghost cell) 
values that have the same order of accuracy as the solution.  
More effort is required to ensure that Neumann boundary conditions do not
reduce the order of accuracy of the solution.  In general, Neumann boundary
conditions require that the accuracy of ghost cells slightly exceed the 
desired order of accuracy for the solution.  A detailed discussion of 
methods for imposing boundary conditions is beyond the scope of this paper.  
We refer interested readers to the thorough analysis presented
in~\cite{gko_book}.  To avoid complicating the discussion in 
Sections~\ref{sec:applications_1d} and~\ref{sec:applications_multidim}, 
Neumann boundary conditions are imposed by taking advantage of knowledge of 
the analytical solution.

@@

\subsubsection{Irregular Domains}
All of the previous examples share an important feature: they are all
defined on simple, rectangular domains.  Thus, any of several methods 
could have been used to obtain solutions with similar or even higher accuracy.
For example, a fourth-order accurate solution to the diffusion equation
(in any number of space dimensions) could be obtained by using a 
pseudospectral spatial discretization~\cite{trefethen_spectral_book, 
boyd_spectral_book} of the Laplacian and a fourth-order time integration scheme 
(\eg Runge-Kutta).  When the domain is irregularly shaped, high-order methods 
are harder to design and implement.  For instance, spectral and pseudospectral 
methods are not directly applicable because the domain is not a simple 
cartesian product of intervals.  While it is possible to use spectral methods
for problems on irregular domains through the clever introduction of a phase 
field~\cite{bueno_orovio_2006, feng_2006}, this approach does not 
actually achieve spectral accuracy because the error is dominated by the 
width of the smoothed approximation to the phase field (as opposed to the 
accuracy of the discretization for the spatial 
derivatives)~\cite{bueno_orovio_2006}.

@@

% FIGURE SHOWING GHOST CELLS AT IRREGULAR BOUNDARY
% PLACED HERE TO SPREAD FIGURES OUT
\begin{figure}[tb]
\begin{center}
\scalebox{0.25}{\includegraphics{figures/ghostcell_edge}} 
\ \ \ \ \ \ \ \ \ \ \ \ \ 
\scalebox{0.25}{\includegraphics{figures/ghostcell_corner}}
\caption{Illustrations of edge (left) and corner (right) ghostcells.
The edge ghostcell, $u_E$, is filled by using cubic extrapolation of 
the value on the boundary, $u_B$, and values at three interior grid points.  
The corner ghostcell, $u_C$, is filled by using a fourth-order accurate
Taylor series expansion centered at $u_{0,0}$, which uses the values of $u$ 
at the specified interior grid points and neighboring edge ghost cells to 
approximate the partial derivatives in the expansion. 
}
\label{fig:ghost_cells}
\end{center}
\end{figure}

@@

@@

% FIGURE SHOWING SOLUTION OF 2D DIFFUSION EQUATION ON STARFISH DOMAIN
% PLACED HERE TO SPREAD FIGURES OUT
\begin{figure}[tb]
\begin{center}
\scalebox{0.35}{
  \includegraphics{figures/diffusion_eqn_2d_starfish_domain_error_vs_N}} 
\caption{$L^\infty$ error as a function of number of grid points for two
finite difference schemes that solve the 2D diffusion equation on an 
irregular domain: forward Euler with optimal time step and correction 
terms (circles) and forward Euler with suboptimal time step $\dt = \dx^2/8 D$
(squares).  Both methods solve the 2D diffusion equation described in the 
caption for Figure~\ref{fig:diffusion_eqn_2d_starfish_domain}.
}
\label{fig:diffusion_eqn_2d_starfish_error}
\end{center}
\end{figure}

@@

\bibliography{OTSPDE}
